Simulation of an Attic Fire in a Wood Frame Residential Structure - Chicago, IL

### 2015-04-29 Studies/2012_Chicago/Report/response.tex

\documentclass[12pt]{article}
\frenchspacing
\begin{document}

\section{Comments from the Editor}
\begin{itemize}
\item We would appreciate more efforts in assessing or estimation somehow the error or uncertainty of your results.
\item An effort to read some of the work published in Fire Technology on the topics at hand would be appreciated by the editorial board.
\item There are too many figures (20); please consider removing some or creating an Appendix where the 5 least important figures are presented.
\item Please consider to submit the input file/s for this work as Electronic Supplementary Material.
\end{itemize}

\section{Comments from Reviewer 1}
\begin{itemize}
\item It may be better to summarize the account in the author's own words, and to link it more closely to Table 1. This would be particularly helpful for section 3.4, since the ventilation conditions are not explained in great detail. \\
{\it The authors were able to view portions of video taken during the incident that allowed for the estimation of failure times and failure areas of exterior walls and windows. This is addressed in the text in Section 3.4. Since this case deals with the death of a firefighter the authors decided to leave the decription of events to narrative from the official NIOSH report.}
\item I miss, however, a more critical point of view that would be essential in guiding fire modellers as to what the limits of applicability of these models is. In this context I would require that the authors elaborate on how they approached the modelling of under-ventilated conditions. FDS is known for running into problems when flaming in vitiated air occurs. In figure 7 the authors present external flaming out of the rear window. Is this realistic? FDS tends to present combustion at vents, even when the flames in the interior have already been extinguished due to lack of oxygen. Is the flow of hot gases into the hallway after the door breakage accompanied by flaming combustion?
\item The authors discard the secondary simulation based on the similarity of the overall HRR, but an alternative ventilation pattern could potentially alter the flow path inside the building. They should explain in more detail why this is not the case in this scenario (or provide images of the flow path and temperature distributions of the secondary simulation).
\item I further think that the beginning of section 5.2 is misleading. Surely second-degree burn injuries are only produced by direct contact with objects at 55C. Air at 55C would not burn the skin. It would be useful to relate the temperatures (gas and wall) to heat fluxes, as this would be a more correct approach in terms of human perception and injuries. \\
{\it The section has been updated.}
\end{itemize}

\section{Comments from Reviewer 2}
\begin{itemize}
 \item The authors mention that the fire was under-ventilated yet the species that they have used in the analysis are well ventilated values.  Although the results of the modelling will not be effected by the assumed fuel, the precedent set in this paper could be used by others to justify these as appropriate values for under-ventilated fire when clearly they are not.  Additional discussion on the species and why the well ventilated values used are appropriate for this analysis but should not be used when modelling under-ventilated fire.
\item The authors used 50kW/m2 for calculating there maximum heat release rate yet this value is lower than expected.  The author should explain why this low value was chosen.  The authors then go on and use a HRRPUA=2250kW/m2.  Why did not use the original 50kW/m2 and expand the area burning?  Additional explanation would be helpful.
\item This results in a floor area of approximately 105 m2 (1100 ft2) per floor and a main living area of approximately 110 m2 (2200 ft2).
The 110m2 should be 210m2??? \\
{\it Thanks for the keen eye. Typo is corrected.}
\item  Around the time of the incident inside the building, E49 attic window.  It is not clear if this was an attic window or the window from the 2nd floor enclosed porch where the fire was.  Urban legend has it that a fire can be pushed through a structure with a hoseline.  Some discussion why this was discounted would be helpful to the reader. \\
{\it The authors recognize that many in the fire service believe fire can be pushed with water but disagree with that legend. A properly applied straight stream or smooth bore stream will not push fire. A wide fog nozzle when applied from the exterior can create a pressure boundary condition at the vent and force hot combustion gas to an area of lower pressure. The text in the tactical consideerations section has been update to include discussion that applying water from the exterior can make conditions better for firefighters.}
\item There was a firefighter with the Captain that perished in the fire.  What were his injuries and location if known?  Some discussion about this 2nd firefighter in the area would be useful. \\
{\it The firefighter with the Captain was not seriously injured.}
\item The condition of the door clearly shows significant damage to the metal door yet the temperatures modelled in FDS are only 260°C.   Is this an artefact of Smokeview's colorbar?  What was the temperature in the porch area when the door is thought to failed?\\
{\it The authors set these bounds because 500~$^{\circ}$F is the upper limit for firefighter PPE. There is discussion of this after the first temperature figure. Text was added to include temperatures on the fire side of the door prior to failure.}
\end{itemize}

### 2015-04-29 Studies/2012_Chicago/Report/firetech_chicago.tex
Chicago Fire: update Fire Tech report and reviewer comments

\subsection{Ventilation}
\label{Vents}
The simulations in this study account for changes in the ventilation due to a combination of fire department operations (opening doors) and fire acting on the structure (fire breaching walls, breaking windows, etc). Two different ventilation scenarios were considered in this study. The first scenario involves a baseline simulation in which the ventilation times and ventilation areas represent our best understanding of the incident. The second scenario, or alternative simulation, does not include the firefighter opening the rear door to make entry to the structure and the subsequent exterior left wall fire penetration. The times of the ventilation changes are provided in Table~\ref{tab:vents}. Structure leakage has been shown to be important when modeling enclosure fires~\cite{beal2009}. The open front door at the start of the simulation and the vents (opening size and open time) created during the simulation are significantly larger than the total effective area of leakage of a structure of this type. Therefore, any leakage in the structure would have a negligible impact on the fluid mechanics within the structure and is not included in this study.

@@

Figure~\ref{fig:temp_129s} shows temperatures of at least 260~$^{\circ}$C (500~$^{\circ}$F) in the second floor rear porch and rear portion of the attic. In this figure, the lower temperature bound is set to 25~$^{\circ}$C (77~$^{\circ}$F), and the upper temperature bound is set to 260~$^{\circ}$C (500~$^{\circ}$F); however, these are only limits on the visualized output, and the actual simulated temperatures may exceed the lower or upper bounds. High temperatures, between 80~$^{\circ}$C (185~$^{\circ}$F) and 200~$^{\circ}$C (390~$^{\circ}$F), also extend down to the rear entrance on the first floor. Before the interior door fails (Fig.~\ref{fig:temp_159s}), the inflow of ambient air from the open rear door reduces the temperatures near the interior doorway, but at the same time, the HRR increases because of the additional ventilation, as shown in Fig.~\ref{fig:hrr}. As shown in Fig.~\ref{fig:temp_161s}, after the second floor interior door fails, the high temperature gas that filled the rear porch flows in the direction of lower pressure (i.e., towards the front door). Note that the conditions in the attic do not appear to change.
\begin{figure}[!ht]
\includegraphics[width=.675\textwidth]{../Figures/west_50th_baseline_161}
\input{../Scripts/colorbar_std} 
\input{../Scripts/colorbar_metric}
\caption{FDS simulated temperature contours, 1~s after interior door failure}
\label{fig:temp_161s}
\end{figure}

@@

\subsection{Assessing the Hazard}
\label{hazard}
A person is susceptible to second-degree burn injuries if exposed to temperatures greater than 55~$^{\circ}$C (130~$^{\circ}$F)~\cite{contactburn}. Although firefighters wear protective gear, gear only offers a finite amount of protection. The polycarbonate in a self contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F)~\cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures of 260~$^{\circ}$C (500~$^{\circ}$F)~\cite{nfpa2013standard}. Prolonged exposure to elevated temperatures can result in a significant amount of heat transferred to the firefighter, putting him/her at risk. Exposure of equipment to temperatures of 260~$^{\circ}$C (500~$^{\circ}$F) represents a Class III exposure~\cite{Donnelly2006}. Firefighters are at increased risk levels when encountering Class III exposure conditions for more than 5~minutes~\cite{Donnelly2006}. Therefore, for the analysis of the simulated exposure temperatures, the maximum visualized temperature is set to indicate all values that are greater than or equal to 260~$^{\circ}$C (500~$^{\circ}$F). This means that temperatures visualized at the maximum range in the displayed model output are at a minimum of 260~$^{\circ}$C (500~$^{\circ}$F); the actual simulated temperatures could in fact be greater.

@@

In this fire incident, the thermal degradation of the steel-faced, wood-framed door resulted in a rapid change in the thermal conditions in the hallway, which resulted in highly hazardous conditions. These conditions would be equivalent to or greater than a Class III exposure~\cite{Donnelly2006}. Prediction of the failure mode and failure time of the door is not a reasonable expectation given the large number of unknowns that the firefighters faced on the fire ground.  In this case, the timing of the interior attack appears to have coincided with the failure time of the door. Research has been conducted~\cite{madrzykowski2009fire, kerber2009fire} and continues to be conducted by NIST, Underwriters Laboratories (UL), and others that demonstrates that applying water from the exterior into the fire area of a structure (prior to the start of interior operations) can significantly improve the safety of firefighters by reducing the thermal hazard from the fire and reducing the potential for developing high velocity hot gas flows within the structure.

>>>>

\subsection{Ventilation}
\label{Vents}
The simulations in this study account for changes in the ventilation due to a combination of fire department operations (opening doors) and fire acting on the structure (fire breaching walls, breaking windows, etc). Two different ventilation scenarios were considered in this study. The first scenario involves a baseline simulation in which the ventilation times and ventilation areas represent our best understanding of the incident. The second scenario, or alternative simulation, does not include the firefighter opening the rear door to make entry to the structure and the subsequent exterior left wall fire penetration. The times of the ventilation changes and estimated areas of openings are provided in Table~\ref{tab:vents}.  Window failures and breaches through the structure (timing and area) were estimated from video taken by a neighbor as well as post incident images. Structure leakage has been shown to be important when modeling enclosure fires~\cite{beal2009}. The open front door at the start of the simulation and the vents (opening size and open time) created during the simulation (to best match the actual fire) are significantly larger than the total effective area of leakage of a structure of this type. Therefore, any leakage in the structure would have a negligible impact on the fluid mechanics within the structure and is not included in this study.

@@

Figure~\ref{fig:temp_129s} shows temperatures of at least 260~$^{\circ}$C (500~$^{\circ}$F) in the second floor rear porch and rear portion of the attic. In this figure, the lower temperature bound is set to 25~$^{\circ}$C (77~$^{\circ}$F), and the upper temperature bound is set to 260~$^{\circ}$C (500~$^{\circ}$F); however, these are only limits on the visualized output to reflect the typically firefighter protective gear limits, and the actual simulated temperatures may exceed the lower or upper bounds. High temperatures, between 80~$^{\circ}$C (185~$^{\circ}$F) and 200~$^{\circ}$C (390~$^{\circ}$F), also extend down to the rear entrance on the first floor. Before the interior door fails (Fig.~\ref{fig:temp_159s}), the inflow of ambient air from the open rear door reduces the temperatures near the interior doorway, but at the same time, the HRR increases because of the additional ventilation, as shown in Fig.~\ref{fig:hrr}. As shown in Fig.~\ref{fig:temp_161s}, after the second floor interior door fails, the high temperature gas that filled the rear porch flows in the direction of lower pressure (i.e., towards the front door). Note that the conditions in the attic do not appear to change. While the visualized temperatures are designed to illustrate the hazard to firefighters, temperatures on the fire side of the doorway prior to failure exceeded 500~$^{\circ}$C (900~$^{\circ}$F).
\begin{figure}[!ht]
\includegraphics[width=.675\textwidth]{../Figures/west_50th_baseline_161}
\input{../Scripts/colorbar_std}
\input{../Scripts/colorbar_metric}
\caption{FDS simulated temperature contours, 1~s after interior door failure}
\label{fig:temp_161s}
\end{figure}

@@

\subsection{Assessing the Hazard}
\label{hazard}
A person is susceptible to second-degree burn injuries if skin temperatures exceed 55~$^{\circ}$C (130~$^{\circ}$F)~\cite{contactburn}. Although firefighters wear protective gear, gear only offers a finite amount of protection. The polycarbonate material in the facepiece of a self-contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F)~\cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures up to 260~$^{\circ}$C (500~$^{\circ}$F) or 10~kW/m$^2$~\cite{nfpa2013standard} for less than 2.5~min. Exposure to elevated temperatures can result in a significant amount of heat transferred to the firefighter, putting him or her at risk.

@@

In this fire incident, the thermal degradation of the steel-faced, wood-framed door resulted in a rapid change in the thermal conditions in the hallway, which resulted in highly hazardous conditions. Prediction of the failure mode and failure time of the door is not a reasonable expectation given the large number of unknowns that the firefighters faced on the fire ground. In this case, the timing of the interior attack appears to have coincided with the failure time of the door.Ongoing research by NIST, Underwriters Laboratories (UL), and others has demonstrated that applying water from the exterior into the fire area of a structure (prior to the start of interior operations) can significantly improve the safety of firefighters by reducing the thermal hazard from the fire and reducing the potential for developing high velocity hot gas flows within the structure~\cite{madrzykowski2009fire, kerber2009fire}.
